---
title: "Examining Physical Activity Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(knitr)
```

```{r, results = 'markup'} 
activity_file = "activity.csv"
act_data = read.csv(activity_file)
summary(act_data)
```

### Histogram of total steps per day
```{r, results = 'markup'}
steps_each_day <- aggregate(act_data$steps, list(act_data$date), FUN = sum)
colnames(steps_each_day) <- c("date","steps")
hist(steps_each_day$steps, breaks = 10, xlab = "Number of steps per day", main="Histogram of steps per day")
```
```{r, echo=FALSE, results='markup'}
png("hist1.png")
hist(steps_each_day$steps, breaks = 10, xlab = "Number of steps per day", main="Histogram of steps per day")
dev.off()
```
### Mean number of steps per day
```{r, results = 'markup'}
mean_steps_per_day <- mean(steps_each_day$steps, na.rm = TRUE)
mean_steps_per_day
```
### Median number of steps per day
```{r, results = 'markup'}
median_steps_per_day<-median(steps_each_day$steps, na.rm = TRUE)
median_steps_per_day
```
### Daily activity pattern
```{r, results = 'markup'}
steps_each_interval <- aggregate(act_data$steps, list(act_data$interval), FUN = mean, na.rm = TRUE)
colnames(steps_each_interval) <- c("interval","steps")
plot(steps_each_interval$interval, steps_each_interval$steps, type = "l", xlab = "Interval", ylab = "Average Number of Steps", col = "blue")
```
```{r, echo=FALSE, results='markup'}
png("daily_pattern.png")
plot(steps_each_interval$interval, steps_each_interval$steps, type = "l", xlab = "Interval", ylab = "Average Number of Steps", col = "blue")
dev.off()
```
### Interval with greatest average number of steps:
```{r, results = 'markup'}
steps_each_interval[which(steps_each_interval$steps == max(steps_each_interval$steps)),1]
```
### Number of missing values in dataset
```{r, results = 'markup'}
length(act_data[which(is.na(act_data$steps)==TRUE),1])
```
### Fill in missing values with average value for that interval
```{r, results = 'markup'}
# These are the records without NA.
act_data_complete <- act_data[which(is.na(act_data$steps)==FALSE),]

# These are the records with NA.
# Use the average steps by interval calculated above to fill in the NAs.
act_data_filled <- merge(act_data[which(is.na(act_data$steps)==TRUE),], 
                         steps_each_interval, by.x="interval", by.y = "interval", all=TRUE)
act_data_filled <- act_data_filled[,c(1,3,4)]
names(act_data_filled)[3] <- "steps"

# combine the two pieces
act_data_revised <- rbind(act_data_complete, act_data_filled)

```
### Histogram of total steps per day using revised data
```{r, results = 'markup'}
steps_each_day_revised <- aggregate(act_data_revised$steps, list(act_data$date), FUN = sum)
colnames(steps_each_day_revised) <- c("date","steps")
hist(steps_each_day_revised$steps, breaks = 10, xlab = "Number of steps each day", main = "Histogram of steps per day")
```
```{r, echo=FALSE, results='markup'}
png("hist2.png")
hist(steps_each_day_revised$steps, breaks = 10, xlab = "Number of steps each day", main = "Histogram of steps per day")
dev.off()
```
### Mean Number of steps per day using revised data
```{r, results = 'markup'}
mean_steps_per_day_revised <- mean(steps_each_day_revised$steps)
mean_steps_per_day_revised
```
### Median steps per day using revised data
```{r, results = 'markup'}
median_steps_per_day_revised<-median(steps_each_day_revised$steps)
median_steps_per_day_revised
```
### Weekend versus weekday patterns
```{r, results = 'markup'}
act_data_weekdays <- act_data_revised
act_data_weekdays$day <- weekdays(as.Date(as.character(act_data$date)))
act_data_weekdays$day[which(act_data_weekdays$day %in% c("Monday", "Tuesday", "Wednesday","Thursday","Friday"))] <- "Weekday"
act_data_weekdays$day[which(act_data_weekdays$day %in% c("Sunday","Saturday"))] <- "Weekend"
act_data_weekdays$day <- as.factor(act_data_weekdays$day)
ww_ints <- aggregate(act_data_weekdays$steps, list(act_data_weekdays$interval, act_data_weekdays$day), FUN = mean, na.rm = TRUE)
colnames(ww_ints) <- c("interval","day","steps")
ggplot(data=ww_ints) + geom_line(mapping=aes(x=interval, y=steps), color="blue") + facet_wrap(~day, ncol=1, nrow=2)
```
```{r, results = 'hide'}
png("weekend_vs_weekday.png")
ggplot(data=ww_ints) + geom_line(mapping=aes(x=interval, y=steps), color="blue") + facet_wrap(~day, ncol=1, nrow=2)
dev.off()
```